Data Prep

### 2024-01-25 00.0_dataPrep.qmd
updated to new who's who, and added a new season of data

```{r download-inpa}
#| eval: false
inpa <- move::getMovebankData(study = 6071688, login = MB.LoginObject, removeDuplicatedTimestamps = TRUE)
inpa <- methods::as(inpa, "data.frame")
inpa <- inpa %>%
  mutate(dateOnly = lubridate::ymd(substr(timestamp, 1, 10)),
         year = as.numeric(lubridate::year(timestamp))) %>%
  filter(lubridate::ymd(dateOnly) >= lubridate::ymd("2020-09-01"), lubridate::ymd(dateOnly) <= lubridate::ymd("2023-05-15")) # cut this off at the same point as the ornitela data
write_feather(inpa, "data/dataPrep/inpa.feather")

length(unique(inpa$trackId)) # 77

@@

```{r download-ornitela}
#| eval: false
# Ornitela: download data from movebank (just a subset of the times for now)
minDate <- "2020-09-01 00:00"
maxDate <- "2023-05-15 11:59"
ornitela <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
write_feather(ornitela, "data/dataPrep/ornitela.feather")

length(unique(ornitela$trackId)) # 127

@@

# Add the Nili_id
ww <- read_excel("data/raw/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, Movebank_id) %>%
  distinct()

@@

-   `data/raw/whoswho_vultures_20230315_new.xlsx`

@@

```{r load-hospital}
joined <- read_feather("data/dataPrep/joined.feather")
periods_to_remove <- read_excel("data/raw/whoswho_vultures_20230315_new.xlsx", sheet = "periods_to_remove")
```

@@

# Join to the original data
removal_annotated <- joined %>%
  left_join(toRemove, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(removal_annotated) == nrow(joined) #T: good, same number of rows.

@@

```{r remove-capture-cage-getroosts}
#| eval: false
# Identify and remove capture dates using Marta's code --------------------
roosts0 <- get_roosts_df(df = removed_periods, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
save(roosts0, file = "data/dataPrep/roosts0.Rda")
```

@@

```{r attach-age-sex}
removed_captures <- read_feather("data/dataPrep/removed_captures.feather")
age_sex <- read_excel("data/raw/whoswho_vultures_20230315_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, birth_year, sex) %>%
  distinct()
```

```{r}
#| eval: false
with_age_sex <- removed_captures %>%
  dplyr::select(-c("sex")) %>%
  left_join(age_sex, by = "Nili_id")
nrow(with_age_sex) == nrow(removed_captures)
length(unique(with_age_sex$Nili_id)) # 196

@@

```{r clean-data}
#| eval: false
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
cleaned <- vultureUtils::cleanData(dataset = with_age_sex, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) 
length(unique(cleaned$Nili_id)) # 172 individuals
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
cleaned <- cleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly))

@@

```{r split-seasons}
#| eval: false
# Split into Marta's 3 seasons --------------------------------------------
cleaned <- cleaned %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2023-05-15") # remove May 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.

with_seasons <- cleaned %>%
  mutate(age = year - birth_year, #XXX consider moving this up to the age/sex section. Why is it happening here?
         sex = case_when(sex == "f/m" ~ "u",
                         TRUE ~ sex)) %>%
  mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
  group_by(Nili_id, year, month) %>%
  mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
  mutate(age_group = case_when(age == 0 ~ "juv",
                               age >= 1 & age <= 4 ~ "sub",
                               age >= 5 ~ "adult",
                               TRUE ~ NA)) %>%
  mutate(seasonUnique = case_when(season %in% c("fall", "summer") ~ paste(year, season, sep = "_"),
                                  season == "breeding" & month == 12 ~ paste(year + 1, season, sep = "_"),
                                  season == "breeding" & month != 12 ~ paste(year, season, sep = "_"))) %>%
  # explicitly set the factor levels in temporal order
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding")),
         season = factor(season, levels = c("breeding", "summer", "fall")))

@@

# REMOVE SUMMER 2020--we don't need to use it
with_seasons <- with_seasons %>%
  filter(seasonUnique != "2020_summer")
length(unique(with_seasons$seasonUnique))

# Separate the seasons -----------------------------
seasons_list <- with_seasons %>%
  group_by(seasonUnique) %>%
  group_split(.keep = T)
length(seasons_list) # should be 8
# extract the season names in case we need a separate vector at some point.
season_names <- map_chr(seasons_list, ~as.character(.x$seasonUnique[1])) # ok good, these are in the right order
save(seasons_list, file = "data/dataPrep/seasons_list.Rda")
save(season_names, file = "data/dataPrep/season_names.Rda")

@@

@@

```{r}
#| eval: false
# Optionally remove individuals with a low fix rate -----------------------
# function to calculate mode (most common) fix rate
Mode <- function(x) { 
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

@@

@@

  .x <- .x %>%
    dplyr::mutate(dateOnly = lubridate::ymd(dateOnly)) %>%
    dplyr::left_join(times, by = "dateOnly") %>%
    dplyr::mutate(daylight = ifelse(timestamp >= sunrise & timestamp <= sunset, "day", "night")) %>%
    dplyr::select(-c(sunrise, sunset)) %>%
    dplyr::filter(daylight == "day")
})

@@

```{r}
#| eval: false
# Calculate number of points per day for each individual
ppd <- map_dfr(removed_northern, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T),
                           meanBatt = mean(battery_charge_percent, na.rm = T),
                           medBatt = mean(battery_charge_percent, na.rm = T)))

>>>>

```{r download-inpa}
inpa <- move::getMovebankData(study = 6071688, login = MB.LoginObject, removeDuplicatedTimestamps = TRUE)
inpa <- methods::as(inpa, "data.frame")
inpa <- inpa %>%
  mutate(dateOnly = lubridate::ymd(substr(timestamp, 1, 10)),
         year = as.numeric(lubridate::year(timestamp))) %>%
  filter(lubridate::ymd(dateOnly) >= lubridate::ymd("2020-09-01"), lubridate::ymd(dateOnly) <= lubridate::ymd("2023-09-15")) # cut this off at the same point as the ornitela data
write_feather(inpa, "data/dataPrep/inpa.feather")

length(unique(inpa$trackId))

@@

```{r download-ornitela}
# Ornitela: download data from movebank (just a subset of the times for now)
minDate <- "2020-09-01 00:00"
maxDate <- "2023-09-15 11:59"
ornitela <- vultureUtils::downloadVultures(loginObject = MB.LoginObject, removeDup = T, dfConvert = T, quiet = T, dateTimeStartUTC = minDate, dateTimeEndUTC = maxDate)
write_feather(ornitela, "data/dataPrep/ornitela.feather")

length(unique(ornitela$trackId))

@@

# Add the Nili_id
ww <- read_excel("data/raw/whoswho_vultures_20230920_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, Movebank_id) %>%
  distinct()

@@

-   `data/raw/whoswho_vultures_20230920_new.xlsx`

@@

```{r load-hospital}
joined <- read_feather("data/dataPrep/joined.feather")
periods_to_remove <- read_excel("data/raw/whoswho_vultures_20230920_new.xlsx", sheet = "periods_to_remove")
```

@@

# Join to the original data
removal_annotated <- joined %>%
  left_join(periods_to_remove, by = c("Nili_id", "dateOnly")) %>%
  mutate(status = replace_na(status, "valid"))
nrow(removal_annotated) == nrow(joined) #T: good, same number of rows.

@@

```{r remove-capture-cage-getroosts}
#| eval: false
# Identify and remove capture dates using Marta's code --------------------
roosts0 <- get_roosts_df(df = removed_periods, id = "Nili_id", timestamp = "timestamp", x = "location_long", y = "location_lat", ground_speed = "ground_speed", speed_units = "m/s", quiet = F)
save(roosts0, file = "data/dataPrep/roosts0.Rda")

# tidy up
rm(roosts0)
```

@@

```{r attach-age-sex}
#| message: false
removed_captures <- read_feather("data/dataPrep/removed_captures.feather")
age_sex <- read_excel("data/raw/whoswho_vultures_20230920_new.xlsx", sheet = "all gps tags")[,1:35] %>%
  dplyr::select(Nili_id, birth_year, sex) %>%
  distinct()
```

```{r}
#| eval: false
with_age_sex <- removed_captures %>%
  dplyr::select(-c("sex")) %>%
  left_join(age_sex, by = "Nili_id")
nrow(with_age_sex) == nrow(removed_captures) # should be TRUE
length(unique(with_age_sex$Nili_id)) #201

@@

```{r clean-data}
#| eval: false
# Clean the data
## Region masking, downsampling, removal of speed outliers, setting altitude outliers to NA, etc.
cleaned <- vultureUtils::cleanData(dataset = with_age_sex, mask = mask, inMaskThreshold = 30, removeVars = F, idCol = "Nili_id", downsample = F, reMask = T) 
length(unique(cleaned$Nili_id)) # 172 individuals
# Fix time zone so dates make sense ---------------------------------------
## Overwrite the dateOnly column from the new times
cleaned <- cleaned %>%
  mutate(timestampIsrael = lubridate::with_tz(timestamp, tzone = "Israel"),
         dateOnly = lubridate::date(timestampIsrael),
         month = lubridate::month(dateOnly),
         year = lubridate::year(dateOnly)) # XXX consider removing the downsample argument. Also this function is insanely slow and needs to be fixed, good grief.

@@

```{r split-seasons}
#| eval: false
# Split into Marta's 3 seasons --------------------------------------------
cleaned <- cleaned %>%
  mutate(start_breeding = ifelse(month %in% c(1:5),
                                 paste(as.character(year-1), "-", "12", "-", "15", sep = ""),
                                 paste(as.character(year), "-", "12", "-", "15", sep = "")),
         start_summer = ifelse(month == 12,
                               paste(as.character(year + 1), "-", "05", "-", "15", sep = ""),
                               paste(as.character(year), "-", "05", "-", "15", sep = "")),
         start_fall = as.Date(paste(as.character(year), "-", "09", "-", "15", sep = "")),
         start_breeding = as.Date(start_breeding),
         start_summer = as.Date(start_summer),
         season = case_when(
           dateOnly >= start_breeding & dateOnly < start_summer ~ "breeding",
           dateOnly >= start_summer & dateOnly <= start_fall ~ "summer",
           dateOnly >= start_fall & dateOnly < start_breeding ~ "fall")) %>%
  filter(dateOnly != "2023-09-15") # remove September 15 2023, because it turns out the season boundaries are supposed to be non-inclusive and this is the latest one.

with_seasons <- cleaned %>%
  mutate(age = year - birth_year, #XXX consider moving this up to the age/sex section. Why is it happening here?
         sex = case_when(sex == "f/m" ~ "u",
                         TRUE ~ sex)) %>%
  mutate(age = ifelse(season == "breeding" & month %in% c(1, 12), age + 1, age)) %>%
  group_by(Nili_id, year, month) %>%
  mutate(age = ifelse(month == 2, min(age) + 1, age)) %>%
  mutate(age_group = case_when(age == 0 ~ "juv",
                               age >= 1 & age <= 4 ~ "sub",
                               age >= 5 ~ "adult",
                               TRUE ~ NA)) %>%
  mutate(seasonUnique = case_when(season %in% c("fall", "summer") ~ paste(year, season, sep = "_"),
                                  season == "breeding" & month == 12 ~ paste(year + 1, season, sep = "_"),
                                  season == "breeding" & month != 12 ~ paste(year, season, sep = "_"))) %>%
  # explicitly set the factor levels in temporal order
  mutate(seasonUnique = factor(seasonUnique, levels = c("2020_summer", "2020_fall", "2021_breeding", "2021_summer", "2021_fall", "2022_breeding", "2022_summer", "2022_fall", "2023_breeding", "2023_summer")),
         season = factor(season, levels = c("breeding", "summer", "fall")))

@@

# REMOVE SUMMER 2020--we don't need to use it
with_seasons <- with_seasons %>%
  filter(seasonUnique != "2020_summer")
length(unique(with_seasons$seasonUnique)) # should be 9

# Separate the seasons -----------------------------
seasons_list <- with_seasons %>%
  group_by(seasonUnique) %>%
  group_split(.keep = T)
length(seasons_list) # should be 9
# extract the season names in case we need a separate vector at some point.
season_names <- map_chr(seasons_list, ~as.character(.x$seasonUnique[1])) # ok good, these are in the right order
length(season_names) # should also be 9
save(seasons_list, file = "data/dataPrep/seasons_list.Rda")
save(season_names, file = "data/dataPrep/season_names.Rda")

@@

-   `data/dataPrep/seasons_list.Rda`
-   `data/dataPrep/season_names.Rda`

@@

```{r}
#| eval: false
# function to calculate mode (most common) fix rate
Mode <- function(x) { 
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

@@

-   `data/dataPrep/removed_lfr.Rda`

@@

  .x <- .x %>%
    dplyr::mutate(dateOnly = lubridate::ymd(dateOnly)) %>%
    dplyr::left_join(times, by = "dateOnly") %>%
    dplyr::mutate(daylight = ifelse(timestamp >= sunrise & timestamp <= sunset, "day", "night")) %>%
    dplyr::filter(daylight == "day")
})

@@

```{r}
#| eval: false
# Calculate number of points per day for each individual
ppd <- map_dfr(removed_northern, ~.x %>%
                 st_drop_geometry() %>%
                 group_by(seasonUnique, season, Nili_id, dateOnly) %>%
                 summarize(n = n(),
                           minBatt = min(battery_charge_percent, na.rm = T)))
